\chapter{Acknowledgements}
\todo{that's new}

Despite being just a bachelor's thesis there is more to putting letters on paper, this would not be possible without people being supportive, inspiring and caring.

First of all I would like to thank my supervisor Mr. Pierre Sachse for the trust and straightforwardness in handling this bachelor's seminar in times of COVID-19. In a time like this I realized it was even more important that I could write about a topic I am passionate in, I am very grateful that I got the chance to write this thesis in the field of psychology of programming and it proved to me that this is the path that I want to take.

Sebastian, Andi, Bernhard and Gregor thank you for believing in me while creating the open source project Statecharts.NET at Innerspace GmbH over the last year, supporting my research in psychology with the technical implementation.

Though the biggest shout out at work goes to you Rebecca, for making me realize again how much I love teaching programming, pair programming and creating tools that make me love programming. Thanks for being my subject of investigation, it was a pleasure to work with you on Statecharts.NET.

No matter if in German or in English, thank you dad for all the awesome discussions, your historical knowledge, your patience and your proof-reading skills!

My friends know how much I love talking about nerdy stuff, thank you for talking about nerdy stuff as well. So many great ideas originated from these conversations! Without you Daniel, Max, Marlene, Iris, David, Christina and Markus this thesis would miss all the knowledge I gained by talking to you about Statecharts.NET.

Teresa, thanks for listening to my daily dose of statecharts, reminding me when it is enough, being there for me and being you!

\vspace{0.5cm}
\noindent
B.\,M.